### 2016-09-27 Chapters/ContinuumSolvationModels.tex
Update a picture
Luca's comments and corrections are in for the main body of the thesis

One can replace the pure environment and interaction terms with their
classical counterparts and obtain the quantum/classical, possibly
\emph{polarizable}, Hamiltonian for an explicit \acrshort{QM}/\acrshort{MM} model.
This brings about the first important point: the need for statistical
averaging.
Whenever a large number of degrees of freedom is involved, one can
access macroscopic observables of the system by the appropriate
averaging of the microscopically detailed motion over phase space
trajectories or on the appropriate statistical
ensemble.\autocite{Hill1960-ql}
The need for ensemble averages leads us to the following observations:
\begin{enumerate*}[label={\alph*)},font={\color{PMS1797}}]
 \item
   the need for \emph{macroscopic} parameters, absent from the
   microscopic Hamiltonian Eq. \eqref{eq:solution-ham}, in carrying out
   statistical simulations.
 \item
   Chosen a thermodynamic ensemble, the basic energetic quantity is
   accordingly determined. For example, the Gibbs free energy  $G$ is
   intrinsically related to the $(NpT)$ ensemble.
 \item
   Even atomistic simulations will lead to results that are essentially
   averages. The discrete picture of the system has been replaced by
   some continuous distribution function.
\end{enumerate*}

These observations strongly suggest that one can perform the averaging
step \emph{before} embarking into the solution of the quantum mechanical
problem and replace the full Hamiltonian with and \emph{effective}
Hamiltonian in which the environment degrees of freedom are properly
averaged.\autocite{Angyan1992-vo, Tapia1992-pu}
The solute-environment interaction term would then be replaced by
a term only depending on the solute degrees of freedom, represented by
the averaged, continuous response functions of the solvent.
This is the physical setting of the \gls{PCM} and is schematically
depicted in Figure \ref{fig:IEF}.
The solute molecule is enclose in a \emph{cavity} inside the
structureless continuum representing the solvent and the basic energetic
quantity is a \emph{free energy}:\autocite{Tomasi1994-gv,
Amovilli1998-pv, Tomasi2005-vm}
\begin{equation}\label{eq:free-energy}
 G = G_\mathrm{cav} + G_\mathrm{el} + G_\mathrm{dis} + G_\mathrm{rep} + G_\mathrm{Mm}(T).
\end{equation}
The decomposition of the solute free energy takes into account the
different physical origins of the different solute-solvent interactions.
$G_\mathrm{cav}$ is the cavity formation free energy, a constant as long
as the solute geometry does not change.\autocite{Tomasi1994-gv,
Floris1997-tu}
The second term $G_\mathrm{el}$ is the most important one in the
framework of continuum models. It accounts for the mutual electrostatic
polarization of solute and solvent, see Section \ref{sec:IEF}.
The third $G_\mathrm{dis}$ and fourth $G_\mathrm{rep}$ terms account for
dispersion and repulsion interactions,
respectively.\autocite{Amovilli1997-fo, Weijo2010-jy}
Finally, $G_\mathrm{Mm}(T)$ accounts for the collective molecular
motions of the solute-solvent system. It explicitly depends on the
temperature $T$ and can be calculated from the solution partition
function, see \noparcite[ref.][]{Amovilli1998-pv} for further details.

In this thesis, we focused exclusively on the purely electrostatic
component of the interaction energy, completely neglecting the other
terms.
While a balanced description of solvation chemistry certainly demands
a proper treatment of all solute-solvent interactions, especially when
non homogeneous environments are studied,\autocite{Mozgawa2014-ad,
Mozgawa2016-wf} it is also true that the electrostatic component is the
most numerically relevant and the most sensitive to changes in the
system studied and in the level of quantum mechanical description.

\begin{figure}[!h]
  \centering
  \includegraphics[width=.5\textwidth]{IEF}
  \caption[The physical setting of the polarizable continuum model.]{
  The physical setting of the \gls{PCM}. The molecular solute,
  represented by its, possibly quantum mechanical, charge density
  $\rho_\mathrm{i}$ is enclosed in a \emph{cavity} $\Omegai$.
  The boundary of the cavity $\Gamma\equiv\partial\Omegai$ is a
  continuously differentiable 2-manifold.
  We assume the material properties of the cavity to be those of vacuum,
  hence characterized by the Green's function
  $\Gi=\frac{1}{|\vect{r}-\vect{r}^\prime|}$.
  The cavity is carved out of an infinite, structureless, continuum
  characterized by its Green's function $\Ge$ and fully covering the
  external subdomain $\Omegae$.
  }
  \label{fig:IEF}
\end{figure}

@@

The molecular solute, represented by its charge density
$\rhoi$ is enclosed in a \emph{cavity} $\Omegai$.
The boundary of the cavity $\Gamma\equiv\partial\Omegai$ is a
2-manifold and is assumed to be continuously differentiable, \ie $\mathcal{C}^1$.
The cavity is carved out of an infinite, structureless, continuum
characterized by the bulk properties of the solvent.
The continuum fully spans the external subdomain $\Omegae$, see Fig. \ref{fig:IEF}.
The source charge density is not assumed to be continuous: both
classical point multipoles and quantum mechanical charge densities can
be treated within the model.
However, we assume that its support is entirely within the cavity,
$\Supp(\rhoi) \subseteq \Omegai$. This assumption is, of course, false
for quantum mechanical charge densities. It can however be proven that
the errors induced by this \emph{charge penetration} are not so
severe.\autocite{Chipman2000-us, Cances2001-qs, Cances2001-qn}

@@

Given the partition of Euclidean space $\mathbb{R}^3$ outlined above,
find the solution $u(\vect{r})$ to the following set:
\begin{subequations}\label{eq:transmission}
  \begin{align}[left={\empheqlbrace}]
  \Li u &= \nabla^2 u = -4\pi\rhoi \,\, \text{in}\,\, \Omegai \label{eq:internal} \\
  \Le u &= 0 \,\, \text{in}\,\, \Omegae \label{eq:external} \\
  [u] &= \ue - \ui = 0 \,\, \text{on}\,\, \Gamma
  \label{eq:trace-jump} \\
[\partial_L u] &= \partiale u - \partiali u = 0 \,\, \text{on}\,\, \Gamma \label{eq:conormal-jump} \\
|u(\vect{r})| &\leq C \|\vect{r} \|^{-1} \,\,\text{for}\,\,\| \vect{r} \|\rightarrow\infty
\label{eq:radiation}
\end{align}
\end{subequations}
Eqs. \eqref{eq:trace-jump} and \eqref{eq:conormal-jump}
are the \emph{jump conditions}, expressed in terms of trace operators
for the solution $u$ and its \emph{conormal} derivative.
For notational simplicity, will use the symbols $\partiale$ and $\partiali$
for the latter and only give it in explicit form when needed.
Further mathematical details and technical results on the definition of
function traces and their use in setting up the proper normal and
conormal derivatives can be found in the excellent book by
\citeauthor{Sauter2011-an}.\autocite{Sauter2011-an}.
The fundamental solutions, or Green's functions, for the elliptic
differential operators $\Li$ and $\Le$ will be denoted by $\Gi$ and
$\Ge$, respectively.
As such the problem is formulated in the so-called \emph{strong} form:
the sought-after solution $u$ has to be at least two times continuously
differentiable.
Moreover, the solution is a function supported over $\mathbb{R}^3$ which
poses challenges to the numerical solution of the problem.
We will take a different approach and reformulate it in terms of
\emph{boundary integral operators}, \ie transform the \acsp{PDE} into a
\gls{BIE}.

@@

The solution of the \acrshort{IEF}-\acrshort{PCM} boundary integral equation can
only be achieved analytically for simple cavity geometries, such as
single spheres and ellipsoids.
When more general molecular cavities are employed one has to resort to a
numerical technique: the \gls{BEM}. This is implemented in four steps:
\begin{enumerate*}[label={\alph*)},font={\color{PMS1797}}]
    \item define the molecular cavity
    \item mesh the molecular surface \label{item:mesh}
    \item discretize the integral equation \label{item:discretize} and
    \item solve the linear systems by a suitable numerical technique.
\end{enumerate*}
Many different approaches to these four steps have been presented, both
in the field of the \acrshort{BEM} and in the context of the \acrshort{PCM}.

The molecular cavity is usually defined as a set of atom-centered
interlocking spheres. The radii are chosen from a suitable set, usually
van der Waals radii.\autocite{Bondi1964-dt, Mantina2009-hb}
If one imagines to model a solvent molecule as a rigid sphere,
three types of molecular surfaces can be defined:
\begin{enumerate*}[label={\alph*)},font={\color{PMS1797}}]
  \item
    the union of the interlocking spheres surfaces is the \gls{vdWS},
    Figure \ref{fig:vdWS}
  \item the locus of points defined by the center of the solvent
    spherical probe while rolling over the \acrshort{vdWS} is the
    \gls{SAS}, Figure \ref{fig:SAS}
  \item the locus of points defined by the contact point of the solvent
    spherical probe while rolling over the \acrshort{vdWS} is the \gls{SES}, Figure \ref{fig:SES}.
\end{enumerate*}
\begin{figure}[tb]
 \centering
 \begin{subfigure}{.45\textwidth}
   \centering
   \includegraphics[width=.85\textwidth]{vdWS.png}
   \caption{\acrlong*{vdWS} (\acrshort{vdWS})}
   \label{fig:vdWS}
 \end{subfigure}
 ~
 \begin{subfigure}{.45\textwidth}
   \centering
   \includegraphics[width=.85\textwidth]{SAS.png}
   \caption{\acrlong*{SAS} (\acrshort{SAS})}
   \label{fig:SAS}
 \end{subfigure}

 \begin{subfigure}{.45\textwidth}
   \centering
   \includegraphics[width=.85\textwidth]{SES.png}
   \caption{\acrlong*{SES} (\acrshort{SES})}
   \label{fig:SES}
 \end{subfigure}
 \caption{Molecular surfaces from interlocking sets of atom-centered spheres.
 Reproduced from \noparcite[ref.][]{Harbrecht2011-dk}.}
\end{figure}
Alternative definitions based on the isodensity surface of the molecular solute
have been put forth.\autocite{Foresman1996-wv}

@@

In this thesis we advocate a different theoretical approach: the
variational formulation of classical polarizable models.
Why so? In such a framework the \acrshort{ASC} becomes an additional,
independent, variationally optimizable degree of freedom.
This makes the coupling to, \eg, extended Lagrangian dynamics
trivial.\autocite{Car1985-jw}
It is well known that electrodynamics can be derived as the minimization
of an action functional.\autocite{Jackson1999}
The principle of least action is however formulated in terms of
\emph{fields} and not more easily manageable quantities, such as the
\acrshort{MEP} or the \acrshort{ASC}.
A true energy functional has to fulfill the following
properties:\autocite{Solis2013-ef}
\begin{enumerate*}[label={\alph*)},font={\color{PMS1797}}]
    \item physical extremal points,
    \item equilibrium values are true energies \label{item:equil-prop} and
    \item convexity (positive-definiteness)
\end{enumerate*}
\citeauthor{Allen2001-fp} and \citeauthor{Attard2003-vr} pioneering
attempts resulted in functionals of scalar densities violating
Property \ref{item:equil-prop} and could not thus be used in extended Lagrangian
dynamics.\autocite{Allen2001-fp, Attard2003-vr, Attard2007-lk}

@@

The functionals in Eqs. \eqref{eq:pcm-functional},
\eqref{eq:mm-functional} and \eqref{eq:pcm-mm-functional} all have a
clear physical meaning.
They describe charge distributions interacting (unfavorably) with
themselves and (favorably) with their respective inducing external
fields and constitutes the polarization energy of the medium.
The topic of variational formulations is particularly important in the
field of \acrshort{MD} simulations.\autocite{Jadhao2012-gf, Jadhao2013-ry, Jadhao2013-hs,
Solis2013-ef}
Our focus however is on the quantum mechanical calculations of molecular
properties and the variational formulation is advantageous since it
dispenses us from manipulating cumbersome nonlinear coupling terms in
the Hamiltonian.
Moreover, the stationarity condition Eq.~\ref{eq:pcm-mm-stationarity}
entails the existence of a classical Hellmann--Feynman theorem:
\begin{equation}\label{eq:classical-HF}
  \deriv{U_\mathrm{pol}}{F} = \pderiv{U_\mathrm{pol}}{F}
  + \pderiv{U_\mathrm{pol}}{\sigma}\pderiv{\sigma}{F}
  + \pderiv{U_\mathrm{pol}}{\kappa}\pderiv{\kappa}{F}
  = \pderiv{U_\mathrm{pol}}{F},
\end{equation}
which is paramount in the formulation of molecular properties, see
Chapter \ref{ch:molprop}.
The variational formulation also lends itself to an implementation
in terms of simultaneous optimization.\autocite{Lipparini2011-aj}

>>>>

One can replace the pure environment and interaction terms with their
classical counterparts and obtain the quantum/classical, possibly
\emph{polarizable}, Hamiltonian for an explicit \acrshort{QM}/\acrshort{MM} model.
This brings about the first important point: the need for statistical
averaging.
Whenever a large number of degrees of freedom is involved, one can
access macroscopic observables of the system by the appropriate
averaging of the microscopically detailed motion over phase space
trajectories or on the appropriate statistical
ensemble.\autocite{Hill1960-ql}
The need for ensemble averages leads us to the following observations:
\begin{enumerate}%[label={\alph*)},font={\color{PMS1797}}]
 \item
   the need for \emph{macroscopic} parameters, absent from the
   microscopic Hamiltonian Eq. \eqref{eq:solution-ham}, in carrying out
   statistical simulations.
 \item
   Chosen a thermodynamic ensemble, the basic energetic quantity is
   accordingly determined. For example, the Gibbs free energy  $G$ is
   intrinsically related to the $(NpT)$ ensemble.
 \item
   Even atomistic simulations will lead to results that are essentially
   averages. The discrete picture of the system has been replaced by
   some continuous distribution function.
\end{enumerate}

These observations strongly suggest that one can perform the averaging
step \emph{before} embarking into the solution of the quantum mechanical
problem and replace the full Hamiltonian with and \emph{effective}
Hamiltonian in which the environment degrees of freedom are properly
averaged.\autocite{Angyan1992-vo, Tapia1992-pu}
The solute-environment interaction term would then be replaced by
a term only depending on the solute degrees of freedom, represented by
the averaged, continuous response functions of the solvent.
This is the physical setting of the \gls{PCM} and is schematically
depicted in Figure \ref{fig:IEF}.
The solute molecule is enclosed in a \emph{cavity} inside the
structureless continuum representing the solvent and the basic energetic
quantity is a \emph{free energy}:\autocite{Tomasi1994-gv,
Amovilli1998-pv, Tomasi2005-vm}
\begin{equation}\label{eq:free-energy}
 G = G_\mathrm{cav} + G_\mathrm{el} + G_\mathrm{dis} + G_\mathrm{rep} + G_\mathrm{Mm}(T).
\end{equation}
The decomposition of the solute free energy takes into account the
different physical origins of the different solute-solvent interactions.
$G_\mathrm{cav}$ is the cavity formation free energy, a constant as long
as the solute geometry does not change.\autocite{Tomasi1994-gv,
Floris1997-tu}
The second term $G_\mathrm{el}$ is the most important one in the
framework of continuum models. It accounts for the mutual electrostatic
polarization of solute and solvent (Section \ref{sec:IEF})
The third and fourth terms account for
dispersion ($G_\mathrm{dis}$) and repulsion ($G_\mathrm{rep}$) interactions,
respectively.\autocite{Amovilli1997-fo, Weijo2010-jy}
Finally, $G_\mathrm{Mm}(T)$ accounts for the collective molecular
motions of the solute-solvent system. It explicitly depends on the
temperature $T$ and can be calculated from the solution partition
function, see \noparcite[ref.][]{Amovilli1998-pv} for further details.

In this thesis, we focused exclusively on the purely electrostatic
component of the interaction energy, completely neglecting the other
terms.
While a balanced description of solvation chemistry certainly demands
a proper treatment of all solute-solvent interactions, especially when
non homogeneous environments are studied,\autocite{Mozgawa2014-ad,
Mozgawa2016-wf} it is also true that the electrostatic component is the
most numerically relevant, especially when considering energy
differences between different electronic states.
%and the most sensitive to changes in the
%system studied and in the level of quantum mechanical description.

\begin{figure}[!h]
  \centering
  \includegraphics[width=.5\textwidth]{IEF.png}
  \caption[The physical setting of the polarizable continuum model.]{
  The physical setting of the \gls{PCM}. The molecular solute,
  represented by its, possibly quantum mechanical, charge density
  $\rho_\mathrm{i}$ is enclosed in a \emph{cavity} $\Omegai$.
  The boundary of the cavity $\Gamma\equiv\partial\Omegai$ is a
  continuously differentiable 2-manifold.
  We assume the material properties of the cavity to be those of vacuum,
  hence characterized by the Green's function
  $\Gi=\frac{1}{|\vect{r}-\vect{r}^\prime|}$.
  The cavity is carved out of an infinite, structureless, continuum
  characterized by its Green's function $\Ge$ and fully covering the
  external subdomain $\Omegae$.
  }
  \label{fig:IEF}
\end{figure}

@@

The molecular solute, represented by its charge density
$\rhoi$ is enclosed in a \emph{cavity} $\Omegai$.
The boundary of the cavity $\Gamma\equiv\partial\Omegai$ is a
2-manifold and is assumed to be continuously differentiable, \ie $\mathcal{C}^1$.
The cavity is carved out of an infinite, structureless, continuum
characterized by the bulk properties of the solvent.
The continuum fully spans the external subdomain $\Omegae$, see Fig. \ref{fig:IEF}.
The source charge density is not assumed to be continuous: both
classical point multipoles and quantum mechanical charge densities can
be treated within the model.
However, we assume that its support is entirely within the cavity,
$\Supp(\rhoi) \subseteq \Omegai$. This assumption is, of course, false
for quantum mechanical charge densities. It can however be proven that
the errors induced by this \emph{charge penetration} are not
severe.\autocite{Chipman2000-us, Cances2001-qs, Cances2001-qn}

@@

Given the partition of Euclidean space $\mathbb{R}^3$ outlined above,
find the solution $u(\vect{r})$ to the following set:
\begin{subequations}\label{eq:transmission}
  \begin{align}[left={\empheqlbrace}]
  \Li u &= \nabla^2 u = -4\pi\rhoi \,\, \text{in}\,\, \Omegai \label{eq:internal} \\
  \Le u &= 0 \,\, \text{in}\,\, \Omegae \label{eq:external} \\
  [u] &= \ue - \ui = 0 \,\, \text{on}\,\, \Gamma
  \label{eq:trace-jump} \\
[\partial_L u] &= \partiale u - \partiali u = 0 \,\, \text{on}\,\, \Gamma \label{eq:conormal-jump} \\
|u(\vect{r})| &\leq C \|\vect{r} \|^{-1} \,\,\text{for}\,\,\| \vect{r} \|\rightarrow\infty
\label{eq:radiation}
\end{align}
\end{subequations}
Eqs. \eqref{eq:trace-jump} and \eqref{eq:conormal-jump}
are the \emph{jump conditions}, expressed in terms of trace operators
for the solution $u$ and its \emph{conormal} derivative.
For notational simplicity, we will use the symbols $\partiale$ and $\partiali$
for the latter and only give it in explicit form when needed.
Further mathematical details and technical results on the definition of
function traces and their use in setting up the proper normal and
conormal derivatives can be found in the excellent book by
\citeauthor{Sauter2011-an}.\autocite{Sauter2011-an}.
The fundamental solutions, or Green's functions, for the elliptic
differential operators $\Li$ and $\Le$ will be denoted by $\Gi$ and
$\Ge$, respectively.
As such the problem is formulated in the so-called \emph{strong} form:
the sought-after solution $u$ has to be at least two times continuously
differentiable.
Moreover, the solution is a function supported over $\mathbb{R}^3$ which
poses challenges to the numerical solution of the problem.
We will take a different approach and reformulate it in terms of
\emph{boundary integral operators}, \ie transform the \acsp{PDE} into a
\gls{BIE}.

@@

The solution of the \acrshort{IEF}-\acrshort{PCM} boundary integral equation can
only be achieved analytically for simple cavity geometries, such as
single spheres and ellipsoids.
When more general molecular cavities are employed one has to resort to a
numerical technique: the \gls{BEM}. This is implemented in four steps:
\begin{enumerate}%[label={\alph*)},font={\color{PMS1797}}]
    \item define the molecular cavity
    \item mesh the molecular surface \label{item:mesh}
    \item discretize the integral equation \label{item:discretize} and
    \item solve the linear systems by a suitable numerical technique.
\end{enumerate}
Many different approaches to these four steps have been presented, both
in the field of the \acrshort{BEM} and in the context of the \acrshort{PCM}.

The molecular cavity is usually defined as a set of atom-centered
interlocking spheres. The radii are chosen from a suitable set, usually
van der Waals radii.\autocite{Bondi1964-dt, Mantina2009-hb}
If one imagines to model a solvent molecule as a rigid sphere,
three types of molecular surfaces can be defined:
\begin{enumerate}%[label={\alph*)},font={\color{PMS1797}}]
  \item
    the union of the interlocking spheres surfaces is the \gls{vdWS},
    Figure \ref{fig:vdWS}
  \item the locus of points defined by the center of the solvent
    spherical probe while rolling over the \acrshort{vdWS} is the
    \gls{SAS}, Figure \ref{fig:SAS}
  \item the locus of points defined by the contact point of the solvent
    spherical probe while rolling over the \acrshort{vdWS} is the \gls{SES}, Figure \ref{fig:SES}.
\end{enumerate}
\begin{figure}[tb]
 \centering
 \begin{subfigure}{.45\textwidth}
   \centering
   \includegraphics[width=.85\textwidth]{vdWS.png}
   \caption{\acrlong*{vdWS} (\acrshort{vdWS})}
   \label{fig:vdWS}
 \end{subfigure}
 ~
 \begin{subfigure}{.45\textwidth}
   \centering
   \includegraphics[width=.85\textwidth]{SAS.png}
   \caption{\acrlong*{SAS} (\acrshort{SAS})}
   \label{fig:SAS}
 \end{subfigure}

 \begin{subfigure}{.45\textwidth}
   \centering
   \includegraphics[width=.85\textwidth]{SES.png}
   \caption{\acrlong*{SES} (\acrshort{SES})}
   \label{fig:SES}
 \end{subfigure}
 \caption{Molecular surfaces from interlocking sets of atom-centered spheres.
 Reproduced from \noparcite[ref.][]{Harbrecht2011-dk}.}
\end{figure}
Alternative definitions based on the isodensity surface of the molecular solute
have also been put forth.\autocite{Foresman1996-wv}

@@

In this thesis we advocate a different theoretical approach: the
variational formulation of classical polarizable models.
In such a framework the \acrshort{ASC} becomes an additional,
independent, variationally optimizable degree of freedom.
This makes the coupling to, \eg, extended Lagrangian dynamics
trivial.\autocite{Car1985-jw}
It is well known that electrodynamics can be derived as the minimization
of an action functional.\autocite{Jackson1999}
The principle of least action is however formulated in terms of
\emph{fields} and not more easily manageable quantities, such as the
\acrshort{MEP} or the \acrshort{ASC}.
A true energy functional has to fulfill the following
properties:\autocite{Solis2013-ef}
\begin{enumerate}%[label={\alph*)},font={\color{PMS1797}}]
    \item physical extremal points,
    \item equilibrium values are true energies \label{item:equil-prop} and
    \item convexity (positive-definiteness)
\end{enumerate}
\citeauthor{Allen2001-fp} and \citeauthor{Attard2003-vr} pioneering
attempts resulted in functionals of scalar densities violating
Property \ref{item:equil-prop} and could not thus be used in extended Lagrangian
dynamics.\autocite{Allen2001-fp, Attard2003-vr, Attard2007-lk}

@@

The functionals in Eqs. \eqref{eq:pcm-functional},
\eqref{eq:mm-functional} and \eqref{eq:pcm-mm-functional} all have a
clear physical meaning.
They describe charge distributions interacting (unfavorably) with
themselves and (favorably) with their respective inducing external
fields and constitutes the polarization energy of the medium.
The topic of variational formulations is particularly important in the
field of \acrshort{MD} simulations.\autocite{Jadhao2012-gf, Jadhao2013-ry, Jadhao2013-hs,
Solis2013-ef}
Our focus is however on the quantum mechanical calculations of molecular
properties and the variational formulation is advantageous since it
dispenses us from manipulating cumbersome nonlinear coupling terms in
the Hamiltonian.
Moreover, the stationarity condition Eq.~\ref{eq:pcm-mm-stationarity}
entails the existence of a classical Hellmann--Feynman theorem:
\begin{equation}\label{eq:classical-HF}
  \deriv{U_\mathrm{pol}}{F} = \pderiv{U_\mathrm{pol}}{F}
  + \pderiv{U_\mathrm{pol}}{\sigma}\pderiv{\sigma}{F}
  + \pderiv{U_\mathrm{pol}}{\kappa}\pderiv{\kappa}{F}
  = \pderiv{U_\mathrm{pol}}{F},
\end{equation}
which is paramount in the formulation of molecular properties, see
Chapter \ref{ch:molprop}.
The variational formulation also lends itself to an implementation
as a simultaneous optimization of the electronic structure and
polarization degrees of freedom.\autocite{Lipparini2011-aj}
